Water desalination using polyelectrolyte hydrogel: Gibbs ensemble modeling

### 2022-09-08 gibbs.tex
started MC sampling in open system.

\subsubsection{Monte Carlo sampling}

General scheme of the Monte Carlo in the \emph{open system} system box with the gel and the monovalent co- and counter-ions ions, 
$\cl$ and $\na$ is allowed to exchange particles with infinite reservoir with constant chemical potential.

>>>>

\subsubsection{Monte Carlo sampling in open system}

Monte Carlo scheme for sampling the exchange of ions in \emph{open system} is based on the formula for free energy of grand canonical ensemble.

\begin{equation}
    \Omega=E-TS+\sum_i\mu_i N_i\label{eq:Omega-GC}
\end{equation}
where $E$ is internal energy, $T$~--- temperature, $S$~--- entropy, $N_i$ is the number of ions of type $i\in\{\na,\ \cl\}$.

%Our model system represents a finite volume $V$ which exchanges particles with the bath of infinite volume.
%The infinity of the volume means, that the particle exchange between the system and the bath does not influence on the chemical potentials in the bath.
%Such ensemble is called grand canonical ensemble; the free energy in this ensemble, the Landau potential, is
The entropy $S$ is given by the Boltzmann formula~\cite{Nagle2004}:
\begin{equation}
    S=\kB\sum_i\ln\frac{V^{N_i}}{N_i!}\label{eq:entropy}
\end{equation}
%\todo{for partiles without interactions}
It accounts for two contributions:
\begin{enumerate}
    \item the combinatorial entropy $S_{c}=-\sum_i\kB\ln N_i!$ which reflects the freedom of choice among $N$ particles
    \item the mixing entropy $S_{m}=\kB \sum_i N_i\ln V$ which reflects the freedom to place the chosen particle randomly within the simulation box. 
\end{enumerate}
$V$ is the unitless volume, \ie the volume measured in units of $\sigma^3$. 
Thus \refeq{eq:Omega-GC} turns to 
%\todo{this is not reason, moreover A.10 is already wrong as for units}
\begin{equation}
    \Omega=E-\kT\ln\prod_i\frac{V^N_i}{N_i!}+\sum_i\mu_i N_i
\label{eq: Omega_GC}
\end{equation}
The change of free energy associated with an exchange of ion pair is
\begin{equation}
    \Delta\Omega=\kT\ln\prod_i\left(V^{\xi}\frac{N_i!}{\left(N_i+\xi\right)!}\right)+\xi\mu_i+\Delta E
\label{eq: DeltaG GC}
\end{equation}
where $\xi=+1$ stands for insertion of an ion pair and $\xi=-1$ for deletion.
In general $\xi$ can be any number, but in case when  $\xi = \pm 1$ the previous equation gets simplified 
\begin{equation}
    \Delta\Omega=\kT\ln V^{2\xi} \prod_i \left(N_i+\theta(\xi)\right)^{-\xi}+\xi\mu_i+\Delta E
\label{eq: DeltaG GC_simpl}
\end{equation}
where $\theta$ is the Heaviside function, $\theta(\xi) = 1, \mathtt{if}\  \xi>0$ , $\theta(\xi) = 0, \mathtt{if}\  \xi<0$

The procedure of the Monte Carlo Sampling is the following \cite{Frenlkel2002_book}
\begin{enumerate}
	\item we propose the new configuration of the system by insertion (or deletion) of an ion pair, $\xi=\pm1$
	\item we accept the new configuration if
	\begin{equation}
        \mathcal{R}^{\xi}<e^{\Delta\Omega/\kT}=V^{2\xi} \prod_i \left(N_i+\theta(\xi)\right)^{-\xi}e^{\left(\Delta E+\xi\mu\right)/\kT}\label{eq: GC acceptance}
	\end{equation}
\end{enumerate}
where $\mathcal{R}$ is uniformly distributed random number in range between $0$ and $1$.

\subsection{Monte Carlo sampling in closed system}
In the same way we derive the acceptance criterion for sampling the closed system (Gibbs ensemble)

The free energy of the Gibbs ensemble is a sum of free energies of the gel and the external volume.

\begin{equation}
    \Omega=E\gel-TS\gel + E\out-TS\out \label{eq:Omega-GC}
\end{equation}

The entropy $S$ is given by the Boltzmann formula~\cite{Nagle2004}:
\begin{equation}
    S=\kB\sum_i\ln\frac{V\gel^{N\gel_i}}{N_i!}\label{eq:entropy}
\end{equation}
%\todo{for partiles without interactions}
It accounts for two contributions:
\begin{enumerate}
    \item the combinatorial entropy $S_{c}=-\sum_i\kB\ln N_i!$ which reflects the freedom of choice among $N$ particles
    \item the mixing entropy $S_{m}=\kB \sum_i N_i\ln V$ which reflects the freedom to place the chosen particle randomly within the simulation box. 
\end{enumerate}
$V$ is the unitless volume, \ie the volume measured in units of $\sigma^3$. 
Thus \refeq{eq:Omega-GC} turns to 
%\todo{this is not reason, moreover A.10 is already wrong as for units}
\begin{equation}
    \Omega=E-\kT\ln\prod_i\frac{V^N_i}{N_i!}+\sum_i\mu_i N_i
\label{eq: Omega_GC}
\end{equation}
The change of free energy associated with an exchange of ion pair is
\begin{equation}
    \Delta\Omega=\kT\ln\prod_i\left(V^{\xi}\frac{N_i!}{\left(N_i+\xi\right)!}\right)+\xi\mu_i+\Delta E
\label{eq: DeltaG GC}
\end{equation}
where $\xi=+1$ stands for insertion of an ion pair and $\xi=-1$ for deletion.
In general $\xi$ can be any number, but in case when  $\xi = \pm 1$ the previous equation gets simplified 
\begin{equation}
    \Delta\Omega=\kT\ln V^{2\xi} \prod_i \left(N_i+\theta(\xi)\right)^{-\xi}+\xi\mu_i+\Delta E
\label{eq: DeltaG GC_simpl}
\end{equation}
where $\theta$ is the Heaviside function, $\theta(\xi) = 1, \mathtt{if}\  \xi>0$ , $\theta(\xi) = 0, \mathtt{if}\  \xi<0$

The procedure of the Monte Carlo Sampling is the following \cite{Frenlkel2002_book}
\begin{enumerate}
	\item we propose the new configuration of the system by insertion (or deletion) of an ion pair, $\xi=\pm1$
	\item we accept the new configuration if
	\begin{equation}
        \mathcal{R}^{\xi}<e^{\Delta\Omega/\kT}=V^{2\xi} \prod_i \left(N_i+\theta(\xi)\right)^{-\xi}e^{\left(\Delta E+\xi\mu\right)/\kT}\label{eq: GC acceptance}
	\end{equation}
\end{enumerate}
where $\mathcal{R}$ is uniformly distributed random number in range between $0$ and $1$.

We model the ionization equilibrium by means of sampling the reaction ensemble~\cite{Smith1994, Turner2008a}.
In the reaction ensemble we perform chemical reactions as simultaneous insertion, deletion or identity change of particles according to the reaction equation.
For example we assume the ionization of an acidic unit which proceeds according to the reaction equation
\begin{equation}
    \mathrm{AH}\stackrel{K}{\leftrightarrows}\A+\h\label{eq: acid reaction}
\end{equation}
We perform the forward direction of this reaction
by inserting a $\h$ particle and replacing $\AH$ by $\A$.
We perform the reverse direction of this reaction
by deleting a $\h$ particle and replacing $\A$ by $\AH$.

In order to account for the particles creation and deletion
we need to introduce standard states for individual reacting species.
The creation of a particle then means that the particle is removed from the standard state
and inserted into the studied system, and vice versa for deletion.
Each species in its standard state has the \emph{standard} chemical potential, $\mu^\ominus_i$.
The choice of standard state for each component is arbitrary
and we usually define it as pure component at standard pressure $10^5$ Pa
and concentration $1 \molperl$~\cite{IUPAC}.

Thus we rewrite the formula for free energy \refeq{eq:Omega-GC}
accounting for particles creation and deletion as
\begin{equation}
    \Omega=E-TS+\sum_i\left(\mu_i-\mu^\ominus_i\right) N_i\label{eq:Omega-RE}
\end{equation}
where the last term in \refeq{eq:Omega-RE} stands for the exchange of particles between the studied system and the standard states.

Let's have the reaction extent $\xi = (N_i - N_{i,0})/\nu_i$
where $N_i$, resp. $N_{i,0}$ is the actual, resp. initial number of the $i$-th particle species,
$\nu_i$ is its stoichiometric coefficient.
and $i$ runs over individual species involved
in the chemical reaction \refeq{eq: acid reaction} $i \in (\h,\ \A,\ \AH)$. 
The stoichiometric coefficients are positive for reaction products
and negative for reactants: $\nu_\h=1$, $\nu_\A=1$, $\nu_\AH=-1$.
Therefore we take $\xi = +1$ for the forward reaction direction
and $\xi = -1$ for the reverse reaction direction.
Then we can write the change of the free energy of the studied system during one reaction step as
\begin{equation}
\Delta\Omega_{\mathrm{re}}=\kT\ln\left(\prod_i V^{\nu_i\xi}\frac{N_i!}{\left(N_i+\nu_i\xi\right)!}\right)+
\xi\left(\sum_i\nu_i\mu_i-\sum_i\nu_i\mu_i^\ominus\right)+\Delta E
\label{eq: DeltaG RE one}
\end{equation}
%\begin{multline}
%    \Delta\Omega_{\mathrm{re}}=\kT\ln\left(\prod_i V^{\nu_i\xi}\frac{N_i!}{\left(N_i+\nu_i\xi\right)!}\right)+\\
%    \xi\left(\sum_i\nu_i\mu_i-\sum_i\nu_i\mu_i^\ominus\right)+\Delta E
%\label{eq: DeltaG RE one}
%\end{multline}

We write $\sum_i\nu_i \mu_i = 0$
for the chemical reaction in equilibrium.

The term $\sum_i\nu_i\mu_i^\ominus$ is the standard Gibbs free energy change
of the reaction \refeq{eq: acid reaction}.
which defines the equilibrium reaction constant \cite{Atkins} as
\begin{equation}
    K = e^{-\sum_i\nu_i \mu_i^\ominus / \kT}
\end{equation}

Finally, the \refeq{eq: DeltaG RE one} translates to
\begin{equation}
    \Delta\Omega_{\mathrm{re}}=\kT\ln\left(K^\xi\prod_i V^{\nu_i\xi}\frac{N_i!}{\left(N_i+\nu_i\xi\right)!}\right)+\Delta E
\label{eq: DeltaG RE two}
\end{equation}

\subsection{Ion pairs insertions\label{sec: ion insertions}}
We treat the exchange of mobile ions with the reservoir as simultaneous insertion or deletion
of ion pairs or sets in the simulation box.
We insert/delete either $\na$ and $\cl$ ion pair or the two $\cl$ ions together with one $\ca$ ion.
Thus we model the insertion as:
\begin{eqnarray}
    \na +  \cl &\leftrightarrows& \emptyset\label{eq:nacl} \\
    \ca + 2\cl &\leftrightarrows& \emptyset \label{eq:cacl}
\end{eqnarray}
with the corresponding equilibrium reaction constants, reflecting the free energy change during the insertion
\begin{eqnarray*}
    K_\mathrm{NaCl}     = \exp\left(\left(\mu_\na+ \mu_\cl \right) /\kT\right), \\
    K_{\mathrm{CaCl}_2} = \exp\left(\left(\mu_\ca+2\mu_\cl \right /\kT\right))
\end{eqnarray*}
%$K_\mathrm{NaCl}= \exp\left(\mu_\na+\mu_\cl \right)$, $K_{\mathrm{CaCl}_2} = \exp\left(\mu_\ca+2\mu_\cl \right)$

the Monte Carlo in the \emph{open system} box with the gel and the monovalent co- and counter-ions ions, 
$\cl$ and $\na$ is allowed to exchange particles with infinite reservoir with constant chemical potential.
